---
title: "Expression clustering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1.0 Format expression matrices (non-normalized and normalized matrices)
```{r}
Expr_no_norm.df <- read.table("~/Dropbox/1.Research/Clam_Work/DE_analysis/1.Expr_matrix/Cja_Csa.TPM.not_cross_norm")
Expr_norm.df <- read.table("~/Dropbox/1.Research/Clam_Work/DE_analysis/1.Expr_matrix/Cja_Csa.TPM.TMM.EXPR.matrix")

head(Expr_no_norm.df)
head(Expr_norm.df)

Expr_no_norm.log2.df <- as.data.frame(t(log2(Expr_no_norm.df[,1:24]+1)))
Expr_norm.log2.df <- as.data.frame(t(log2(Expr_norm.df[,1:24]+1)))

Expr_no_norm.log2.df <- Expr_no_norm.log2.df[,colSums(Expr_no_norm.log2.df^2) !=0]
Expr_norm.log2.df <- Expr_norm.log2.df[,colSums(Expr_norm.log2.df^2) !=0]

species <- c(rep("Cja",12), rep("Csa",12))
tissues <- rep(c("Ad","F","G","M"), 6)
```

## 1.05 PCA
```{r}
Expr_no_norm.log2.PCA <- prcomp(Expr_no_norm.log2.df, center=T, scale. = T)
Expr_norm.log2.PCA <- prcomp(Expr_norm.log2.df, center=T, scale. = T)
plot(Expr_no_norm.log2.PCA, type = "l")
plot(Expr_norm.log2.PCA, type = "l")
```

## Print PCA BEFORE normalization
```{r}
library(ggbiplot)
g <- ggbiplot(Expr_no_norm.log2.PCA, choices = 1:2, obs.scale = 1, var.scale = 1, 
              groups = species, ellipse = T, circle = T, var.axes = F, labels = rownames(Expr_no_norm.log2.PCA$x))
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)

g <- ggbiplot(Expr_no_norm.log2.PCA, choices = 2:3, obs.scale = 1, var.scale = 1, 
              groups = tissues, ellipse = T, circle = T, var.axes = F, labels = rownames(Expr_no_norm.log2.PCA$x))
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)
```

## 1.1 Print PCA AFTER normalization
```{r}
library(ggbiplot)
library(grid)
library(gridExtra)
g1 <- ggbiplot(Expr_norm.log2.PCA, choices = 1:2, obs.scale = 1, var.scale = 1, 
              groups = species, ellipse = T, circle = T, var.axes = F, labels = rownames(Expr_norm.log2.PCA$x))
g1 <- g1 + scale_color_discrete(name = '')
g1 <- g1 + theme_bw() + theme(legend.direction = 'horizontal',
                            legend.position = "top") + coord_equal(0.2)

g2 <- ggbiplot(Expr_norm.log2.PCA, choices = 2:3, obs.scale = 1, var.scale = 1, 
              groups = tissues, ellipse = T, circle = T, var.axes = F, labels = rownames(Expr_norm.log2.PCA$x))
g2 <- g2 + scale_color_discrete(name = '')
g2 <- g2 + theme_bw() + theme(legend.direction = 'horizontal', 
               legend.position = 'top') + coord_equal(0.4)
grid.arrange(g1,g2,ncol=1)
```

## 1.2 NMDS (nonmetric multidimensional scaling)
```{r}
library(MASS)
TPM.log2.df <- as.data.frame(t(log2(Expr_norm.df[,1:24]+1)))
TPM.log2.df <- TPM.log2.df[,colSums(TPM.log2.df^2) !=0]
TPM.log2.d <- dist(TPM.log2.df)

# Three dimensions
library(scatterplot3d)
fit <- isoMDS(TPM.log2.d, k=3)
NMDS_plot <- data.frame(matrix(data = row.names(fit$points), nrow = 24))
NMDS_plot$x <- fit$points[,1] ## main coordinate that separates by species
NMDS_plot$y <- fit$points[,2] ## tissue
NMDS_plot$z <- fit$points[,3] ## tissue
NMDS_plot$pcol <- matrix(NA, nrow=24)
NMDS_plot[c(1,5,9,13,17,21),]$pcol = "red"  ## Ad
NMDS_plot[c(2,6,10,14,18,22),]$pcol = "blue" ## F
NMDS_plot[c(3,7,11,15,19,23),]$pcol = "black" ## G
NMDS_plot[c(4,8,12,16,20,24),]$pcol = "green" ## M

s3d <- with(NMDS_plot, scatterplot3d(y,z,x, color = pcol, pch=19, type="h", xlab = "NMDS2", ylab = "NMDS3", zlab = "NMDS1"))
s3d.coords <- s3d$xyz.convert(NMDS_plot$y, NMDS_plot$z, NMDS_plot$x)
text(s3d.coords$x, s3d.coords$y, labels = row.names(TPM.log2.df), cex=.7)
```


## 2.0 zFPKM formula
```{r, echo=FALSE}
z_fpkm<-function(i){
  if(all(i>0)) stop('Input not log2 transformed.')
  if(all(!is.na(i))) stop('0\'s need to be NA\'s.')
  my<-density(i,na.rm=T)$x[which.max(density(i,na.rm=T)$y)]
  U<-mean(i[i>my],na.rm=T)
  sigma<-(U-my)*(.5*pi)^.5
  z<-(i-my)/sigma
  z[z< -3]<-NA
  return(z)
}
```

## 2.1 Format expression matrices (FPKM and zFPKM matrices)
```{r}
FPKM.df <- read.table("~/Dropbox/1.Research/Clam_Work/DE_analysis/1.Expr_matrix/Cja_Csa.FPKM.matrix")
FPKM.df[FPKM.df == 0] <- NA
head(FPKM.df)

FPKM_log2.df <- as.data.frame(log2(FPKM.df[,1:24]))
FPKM_log2_tp.df <- t(FPKM_log2.df)
FPKM_log2_tp.df <- FPKM_log2_tp.df[,colSums(FPKM_log2_tp.df^2) !=0]
idx <- apply(FPKM_log2_tp.df, 2, function(x) all(is.na(x)))
FPKM_log2_tp.df <- FPKM_log2_tp.df[,!idx] ## 6373 genes

zFPKM.df <- apply(FPKM_log2.df, 2, z_fpkm)
zFPKM_tp.df <- t(zFPKM.df)
zFPKM_tp.df <- zFPKM_tp.df[,colSums(zFPKM_tp.df^2) !=0]
idx <- apply(zFPKM_tp.df, 2, function(x) all(is.na(x)))
zFPKM_tp.df <- zFPKM_tp.df[,!idx] ## 6373 expressed genes (at least in one library)

species <- c(rep("Cja",12), rep("Csa",12))
tissues <- rep(c("Ad","F","G","M"), 6)
```

## 2.2 PCA
```{r}
FPKM_log2.PCA <- prcomp(FPKM_log2_tp.df, center=T, scale. = T)
plot(FPKM_log2.PCA, type = "l")

zFPKM.PCA <- prcomp(zFPKM_tp.df, center=T, scale. = T)
plot(zFPKM.PCA, type = "l")
```

## 2.3 Print PCA for log2(FPKM)
```{r}
g1 <- ggbiplot(FPKM_log2.PCA, choices = 1:2, obs.scale = 1, var.scale = 1, 
              groups = species, ellipse = T, circle = T, var.axes = F, labels = rownames(Expr_norm.log2.PCA$x))
g1 <- g1 + scale_color_discrete(name = '')
g1 <- g1 + theme_bw() + theme(legend.direction = 'horizontal',
                            legend.position = "top") + coord_equal(0.4)

g2 <- ggbiplot(FPKM_log2.PCA, choices = 2:3, obs.scale = 1, var.scale = 1, 
              groups = tissues, ellipse = T, circle = T, var.axes = F, labels = rownames(Expr_norm.log2.PCA$x))
g2 <- g2 + scale_color_discrete(name = '')
g2 <- g2 + theme_bw() + theme(legend.direction = 'horizontal', 
               legend.position = 'top') + coord_equal(0.4)
grid.arrange(g1,g2,ncol=1)
```


## 2.4 Print PCA for zFPKM  normalization
```{r}
g1 <- ggbiplot(zFPKM.PCA, choices = 1:2, obs.scale = 1, var.scale = 1, 
              groups = species, ellipse = T, circle = T, var.axes = F, labels = rownames(Expr_norm.log2.PCA$x))
g1 <- g1 + scale_color_discrete(name = '')
g1 <- g1 + theme_bw() + theme(legend.direction = 'horizontal',
                            legend.position = "top") + coord_equal(0.4)

g2 <- ggbiplot(zFPKM.PCA, choices = 2:3, obs.scale = 1, var.scale = 1, 
              groups = tissues, ellipse = T, circle = T, var.axes = F, labels = rownames(Expr_norm.log2.PCA$x))
g2 <- g2 + scale_color_discrete(name = '')
g2 <- g2 + theme_bw() + theme(legend.direction = 'horizontal', 
               legend.position = 'top') + coord_equal(0.4)
grid.arrange(g1,g2,ncol=1)
```



## 2.5 NMDS (nonmetric multidimensional scaling) for log2(FPKM)
```{r}
library(MASS)
FPKM_log2_tp.d <- dist(FPKM_log2_tp.df)

# Three dimensions
library(scatterplot3d)
fit <- isoMDS(FPKM_log2_tp.d, k=3)
NMDS_plot <- data.frame(matrix(data = row.names(fit$points), nrow = 24))
NMDS_plot$x <- fit$points[,1] ## main coordinate that separates by species
NMDS_plot$y <- fit$points[,2] ## tissue
NMDS_plot$z <- fit$points[,3] ## tissue
NMDS_plot$pcol <- matrix(NA, nrow=24)
NMDS_plot[c(1,5,9,13,17,21),]$pcol = "red"  ## Ad
NMDS_plot[c(2,6,10,14,18,22),]$pcol = "blue" ## F
NMDS_plot[c(3,7,11,15,19,23),]$pcol = "black" ## G
NMDS_plot[c(4,8,12,16,20,24),]$pcol = "green" ## M

s3d <- with(NMDS_plot, scatterplot3d(y,z,x, color = pcol, pch=19, type="h", xlab = "NMDS2", ylab = "NMDS3", zlab = "NMDS1"))
s3d.coords <- s3d$xyz.convert(NMDS_plot$y, NMDS_plot$z, NMDS_plot$x)
text(s3d.coords$x, s3d.coords$y, labels = row.names(FPKM_log2_tp.df), cex=.7)
```


## 2.6 NMDS (nonmetric multidimensional scaling) for zFPKM
```{r}
library(MASS)
zFPKM.d <- dist(zFPKM_tp.df)

# Two dimensions
fit <- isoMDS(zFPKM.d, k=2)
NMDS_plot <- data.frame(matrix(data = row.names(fit$points), nrow = 24))
NMDS_plot$x <- fit$points[,1] ## main coordinate that separates by species
NMDS_plot$y <- fit$points[,2] ## tissue
NMDS_plot$pcol <- matrix(NA, nrow=24)
NMDS_plot[c(1,5,9,13,17,21),]$pcol = "red"  ## Ad
NMDS_plot[c(2,6,10,14,18,22),]$pcol = "blue" ## F
NMDS_plot[c(3,7,11,15,19,23),]$pcol = "black" ## G
NMDS_plot[c(4,8,12,16,20,24),]$pcol = "green" ## M

with(NMDS_plot, plot(x, y, xlab="NMDS 1", ylab="NMDS 2", 
  main="zFPKM nonmetric multi-dimensional scaling (NMDS)",	type="n"))
text(NMDS_plot$x, NMDS_plot$y, labels = row.names(zFPKM_tp.df), cex=.7, col = NMDS_plot$pcol)

# Three dimensions
library(scatterplot3d)
fit <- isoMDS(zFPKM.d, k=3)
NMDS_plot <- data.frame(matrix(data = row.names(fit$points), nrow = 24))
NMDS_plot$x <- fit$points[,1] ## main coordinate that separates by species
NMDS_plot$y <- fit$points[,2] ## tissue
NMDS_plot$z <- fit$points[,3] ## tissue
NMDS_plot$pcol <- matrix(NA, nrow=24)
NMDS_plot[c(1,5,9,13,17,21),]$pcol = "red"  ## Ad
NMDS_plot[c(2,6,10,14,18,22),]$pcol = "blue" ## F
NMDS_plot[c(3,7,11,15,19,23),]$pcol = "black" ## G
NMDS_plot[c(4,8,12,16,20,24),]$pcol = "green" ## M

s3d <- with(NMDS_plot, scatterplot3d(y,z,x, color = pcol, pch=19, type="h", xlab = "NMDS2", ylab = "NMDS3", zlab = "NMDS1"))
s3d.coords <- s3d$xyz.convert(NMDS_plot$y, NMDS_plot$z, NMDS_plot$x)
text(s3d.coords$x, s3d.coords$y, labels = row.names(zFPKM_tp.df), cex=.7)
```

